### 2009-05-22 paper.tex
small edits

Due to this possibility of unobserved ancestral sequence,
the standard CYK and Inside algorithms fail on the ensemble model.
A potentially-unbounded number of residues in the ancestor can be
absent in all extant sequences.
This manifests as cycles of $\Snull$ states within the multiple-sequence SCFG,
states which emit to the ancestral sequence but not to any extant sequence.
The presence of bifurcations with one or more possibly-empty child branches,
corresponding to entire substructures in the ancestral sequence that
are absent in any extant sequence, further complicates inference.
While there exists a straightforward fix for maximum-likelihood
inference with the CYK algorithm, the Inside algorithm requires
further computational machinery.

In order to correctly perform maximum-likelihood inference, we only
require that the maximum-likelihood parse tree corresponds to the parse
returned by the CYK algorithm; the entire
probability distribution over parses need not be preserved.
Because a state path which includes a cycle of $\Snull$ states
necessarily has lower probability than an equivalent one without, we
can simply ignore such $\Snull$ cycles during CYK.
Bifurcations with empty child branches cannot be similarly ignored,
but this problem can be addressed by creating equivalent effective
direct transitions between the parents of these bifurcation states and
their non-empty children.  Once these transformations have been performed,
maximum-likelihood inference can be performed using a standard
three-sequence CYK algorithm on the ensemble model.
(Similarly, the null cycle elimination discussed in computational
linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)
However, posterior-decoding inference, which uses the Inside and
Outside algorithms, required a modified DP algorithm which preserves
the entire probability distribution over parses; the transformations
just described to make CYK work properly are insufficient for the
Inside and Outside algorithms.

There are two possible solutions to this difficulty: We can either
create new versions of the standard inference algorithms for SCFGs
which can cope with SCFGs with null cycles or we can modify the
structure of the SCFGs generated by our model-construction algorithm
such that they are free of null cycles.  The first approach of
creating new algorithms is simple in principle---we can simply
iteratively re-fill entries of the DP matrix, corresponding to summing
the probabilities for unobserved null cycles, until the approximation
is sufficiently good---but it is difficult to implement efficiently
and necessarily has worse time complexity than the standard inference
algorithms due to the need to iterate over parts of the (large) DP
matrix.  The second approach, of modifying the structure of the SCFGs
themselves, has a closed-form solution for the case of HMMs, but to our
knowledge, efficient exact null-cycle elimination for SCFGs is an open
mathematical problem (complicated by the presence of null cycles
involving bifurcations).

@@

The ensemble grammar contains many rules that can be applied redundantly,
together or in isolation, to generate subtrees of the parse tree that do
not generate any terminals, and could be replaced with simpler rules.
In fact, this is a general property of SCFGs with null states and/or bifurcations.
In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules.
We begin by identifying two types of redundant parse-structures: {\bf empty bifurcations} and {\bf empty paths}.

>>>>

Due to this possibility of unobserved ancestral sequence,
the standard CYK and Inside recursions fail on the ensemble model:
that is, they contain cyclic self-references which cannot be resolved.
The intuitive reason for this is that a (potentially-unbounded) number of residues in the ancestor can be
absent in all extant sequences.
This manifests as cycles of $\Snull$ states within the multiple-sequence SCFG,
states which emit to the ancestral sequence but not to any extant sequence.
The presence of bifurcations with one or more possibly-empty child branches,
corresponding to entire substructures in the ancestral sequence that
are absent in any extant sequence, further complicates inference.
While there exists a straightforward fix for the CYK algorithm, the Inside algorithm requires more work.

Consider the CYK algorithm first.
In order to correctly perform maximum-likelihood inference, we only
require that the maximum-likelihood parse tree corresponds to the parse
returned by the CYK algorithm; the entire
probability distribution over parses need not be preserved.
Because a state path which includes a cycle of $\Snull$ states
necessarily has lower probability than an equivalent one without, we
can simply ignore such $\Snull$ cycles during CYK.
Bifurcations with empty child branches cannot be similarly ignored,
but this problem can be addressed by creating equivalent effective
direct transitions between the parents of these bifurcation states and
their non-empty children.  Once these transformations have been performed,
maximum-likelihood inference can be performed using a standard
three-sequence CYK algorithm on the ensemble model.
(Similarly, the null cycle elimination discussed in computational
linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)

The transformations just described to make CYK work properly are insufficient for the
Inside and Outside algorithms.
For CYK, null cycles can be ignored, since they should never appear in the maximum likelihood parse.
For Inside and Outside, null cycles cannot be ignored, since they will contribute to the likelihood sum over parse trees.

There are two possible solutions to this difficulty: we can either
create new versions of the standard inference algorithms for SCFGs
which can cope with SCFGs with null cycles or we can modify the
structure of the SCFGs generated by our model-construction algorithm
such that they are free of null cycles.  The first approach of
creating new algorithms is simple in principle---we can simply
iteratively re-fill entries of the DP matrix, corresponding to summing
the probabilities for unobserved null cycles, until the approximation
is sufficiently good---but it is difficult to implement efficiently
and necessarily has worse time complexity than the standard inference
algorithms due to the need to iterate over parts of the (large) DP
matrix.  The second approach, of modifying the structure of the SCFGs
themselves, has a closed-form solution for the case of HMMs, but to our
knowledge, efficient exact null-cycle elimination for SCFGs is an open
problem (complicated by the presence of null cycles
involving bifurcations).

@@

As noted above, the ensemble grammar contains many rules that can be applied redundantly,
together or in isolation, to generate subtrees of the parse tree that do
not generate any terminals, and could be replaced with simpler rules.
In fact, this is a general property of SCFGs with null states and/or bifurcations.
In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules.
We begin by identifying two types of redundant parse-structures: {\bf empty bifurcations} and {\bf empty paths}.
